---
title: "ATRT Heat Maps"
output: 
  html_document:
    fig_width: 14
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ComplexHeatmap, quietly = TRUE)
library(BrainSABER)

```

## Cosine Similarity Heatmaps
```{r MYCcosSim, echo=FALSE}
data("MYCcosMat")
ComplexHeatmap::Heatmap(t(MYCcosMat), cluster_columns = FALSE,       column_names_gp = gpar(fontsize = 8), name = "MYC Group", column_title = "MYC Group", column_title_gp = gpar(fontsize = 20, fontface = "bold"))
```


Gene Vector Used: MYC group at 8pcw
TNFRSF1A      BCL3      NEFH      EZH2      EZH1       MYC     ERBB2 
 4.484885  0.121902  0.510035 22.393495  4.095582  4.840407 12.126491 
     SHC1    COL1A2    COL3A1     MYD88    COL5A2 
 6.374071 25.341298 47.432334 11.592439  7.011703 
 

``` {r SHHcos, echo=FALSE}
ComplexHeatmap::Heatmap(t(SHHcosMat), cluster_columns = FALSE,  column_names_gp = gpar(fontsize = 8), name = "SHH Group", column_title = "SHH Group", column_title_gp = gpar(fontsize = 20, fontface = "bold"))
```

Gene Vector Used: SHH Group at 8pcw
 TCF3       GLI2     SEMA5B     SEMA6A      EFNA2     CRNKL1       CDK6 
 58.955694   4.116634   9.368359   2.798069   4.967494  15.735434   7.169478 
      EZH2       EZH1      AADAT       GLI1       LIFR     ACVR2B       APC2 
 22.393495   4.095582   6.568030   0.470286   7.401808   2.981786   9.824244 
       MLL      GNAI1      RBM8A       MYCN    WHSC1L1     DPYSL5        SHH 
  4.048180  25.523049  24.182789  64.725336   9.290797  51.558048   0.057300 
   HNRNPA3        ID4      ARNT2      SF3A3       PBX1      PTCH1     SRSF10 
 74.747579 310.269520  19.263336  24.377935   9.371658   1.029264  18.521356 
    SRGAP1 
  2.032065 

```{r TYRcos, echo=FALSE}
ComplexHeatmap::Heatmap(t(TYRcosMat), cluster_columns = FALSE, column_names_gp = gpar(fontsize = 8), name = "TYR Group", column_title = "TYR Group",  column_title_gp = gpar(fontsize = 20, fontface = "bold"))
```

Gene Vector Used: TYR Group at 8pcw
    DGAT2      DLG1       TYR      EDN1     PSEN1      TCF7     WNT11     BAMBI 
 0.259309  6.205125  0.024053  0.406282 11.027938  0.269182  0.177531  0.918871 
    LPIN2      EZH2      EZH1     CCND1     VEGFA      WWC1   GALNT12     SFRP5 
 5.434781 22.393495  4.095582 22.889201  4.080633  5.857789  0.727062  0.039025 
    EPHX2       CAT      BMP4      GDF5     ECHS1    NFATC1   HSD17B4     LPIN1 
 1.431250 21.662408  2.011776  0.353537 46.677636  0.141373 20.373535  1.127990 
   GALNT5     NUDT7     ERBB2      WTIP   ALDH9A1    GALNT2     ACSL1     WNT7A 
 0.073839  1.680173 12.126491  2.982206 24.028322 20.360967  2.955874  4.274548 
  GALNT10      FZD6      SOST     ACOX2    AGPAT2    CYP7B1      MITF     OGDHL 
 1.327176  2.036004  0.015721  0.090484  4.531440  1.238006  0.398427  0.056332 

## Euclidean Distance Heatmaps

```{r MYCeucSim, echo=FALSE}
data("MYCeucMat")
ComplexHeatmap::Heatmap(t(MYCeucMat), cluster_columns = FALSE,       column_names_gp = gpar(fontsize = 8), name = "MYC Group", column_title = "MYC Group", column_title_gp = gpar(fontsize = 20, fontface = "bold"))
```


Gene Vector Used: MYC group at 8pcw
TNFRSF1A      BCL3      NEFH      EZH2      EZH1       MYC     ERBB2 
 4.484885  0.121902  0.510035 22.393495  4.095582  4.840407 12.126491 
     SHC1    COL1A2    COL3A1     MYD88    COL5A2 
 6.374071 25.341298 47.432334 11.592439  7.011703 
 

``` {r SHHeuc, echo=FALSE}
ComplexHeatmap::Heatmap(t(SHHeucMat), cluster_columns = FALSE,  column_names_gp = gpar(fontsize = 8), name = "SHH Group", column_title = "SHH Group", column_title_gp = gpar(fontsize = 20, fontface = "bold"))
```

Gene Vector Used: SHH Group at 8pcw
 TCF3       GLI2     SEMA5B     SEMA6A      EFNA2     CRNKL1       CDK6 
 58.955694   4.116634   9.368359   2.798069   4.967494  15.735434   7.169478 
      EZH2       EZH1      AADAT       GLI1       LIFR     ACVR2B       APC2 
 22.393495   4.095582   6.568030   0.470286   7.401808   2.981786   9.824244 
       MLL      GNAI1      RBM8A       MYCN    WHSC1L1     DPYSL5        SHH 
  4.048180  25.523049  24.182789  64.725336   9.290797  51.558048   0.057300 
   HNRNPA3        ID4      ARNT2      SF3A3       PBX1      PTCH1     SRSF10 
 74.747579 310.269520  19.263336  24.377935   9.371658   1.029264  18.521356 
    SRGAP1 
  2.032065 

```{r TYReuc, echo=FALSE}
ComplexHeatmap::Heatmap(t(TYReucMat), cluster_columns = FALSE, column_names_gp = gpar(fontsize = 8), name = "TYR Group", column_title = "TYR Group",  column_title_gp = gpar(fontsize = 20, fontface = "bold"))
```

Gene Vector Used: TYR Group at 8pcw
    DGAT2      DLG1       TYR      EDN1     PSEN1      TCF7     WNT11     BAMBI 
 0.259309  6.205125  0.024053  0.406282 11.027938  0.269182  0.177531  0.918871 
    LPIN2      EZH2      EZH1     CCND1     VEGFA      WWC1   GALNT12     SFRP5 
 5.434781 22.393495  4.095582 22.889201  4.080633  5.857789  0.727062  0.039025 
    EPHX2       CAT      BMP4      GDF5     ECHS1    NFATC1   HSD17B4     LPIN1 
 1.431250 21.662408  2.011776  0.353537 46.677636  0.141373 20.373535  1.127990 
   GALNT5     NUDT7     ERBB2      WTIP   ALDH9A1    GALNT2     ACSL1     WNT7A 
 0.073839  1.680173 12.126491  2.982206 24.028322 20.360967  2.955874  4.274548 
  GALNT10      FZD6      SOST     ACOX2    AGPAT2    CYP7B1      MITF     OGDHL 
 1.327176  2.036004  0.015721  0.090484  4.531440  1.238006  0.398427  0.056332 


## All Groups Together

```{r allGroups, echo=FALSE}
library(ComplexHeatmap, quietly = TRUE)
library(BrainSABER, quietly = TRUE)
data("ATRTcosMat","ATRTeucMat")
ComplexHeatmap::Heatmap(t(ATRTcosMat), cluster_columns = FALSE, row_names_gp = gpar(fontsize = 8), name = "Cosine Similarity", column_title = "Cosine Similarity", column_title_gp = gpar(fontsize = 20, fontface = "bold", na_col = "gray"))

ComplexHeatmap::Heatmap(t(ATRTeucMat), cluster_columns = FALSE, row_names_gp = gpar(fontsize = 8), name = "Euclidean Distance", column_title = "Euclidean Distance", column_title_gp = gpar(fontsize = 20, fontface = "bold", na_col = "gray"))
