---
title: "ATRT Heat Maps"
output: 
  html_document:
    fig_width: 14
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Cosine Similarity Heatmaps
```{r cosSim, echo=FALSE}
library(ComplexHeatmap, quietly = TRUE)
library(BrainSABER, quietly = TRUE)
data("MYCcosMat","SHHcosMat","TYRcosMat")
MYCc <- ComplexHeatmap::Heatmap(t(MYCcosMat), cluster_columns = FALSE,       column_names_gp = gpar(fontsize = 8), name = "MYC Group", column_title = "MYC Group", column_title_gp = gpar(fontsize = 20, fontface = "bold"))
SHHc <- ComplexHeatmap::Heatmap(t(SHHcosMat), cluster_columns = FALSE,  column_names_gp = gpar(fontsize = 8), name = "SHH Group", column_title = "SHH Group", column_title_gp = gpar(fontsize = 20, fontface = "bold"))
TYRc <- ComplexHeatmap::Heatmap(t(TYRcosMat), cluster_columns = FALSE, column_names_gp = gpar(fontsize = 8), name = "TYR Group", column_title = "TYR Group",  column_title_gp = gpar(fontsize = 20, fontface = "bold"))

hlistc <- MYCc+ SHHc + TYRc

ComplexHeatmap::draw(hlistc)
```

## Euclidean Distance Heatmaps

```{r eucSim, echo=FALSE}
library(ComplexHeatmap, quietly = TRUE)
library(BrainSABER, quietly = TRUE)
data("MYCeucMat","SHHeucMat","TYReucMat")
MYCe <- ComplexHeatmap::Heatmap(t(MYCeucMat), cluster_columns = FALSE, column_names_gp = gpar(fontsize = 8), name = "MYC Group", column_title = "MYC Group", column_title_gp = gpar(fontsize = 20, fontface = "bold"))
SHHe <- ComplexHeatmap::Heatmap(t(SHHeucMat), cluster_columns = FALSE,  column_names_gp = gpar(fontsize = 8), name = "SHH Group", column_title = "SHH Group", column_title_gp = gpar(fontsize = 20, fontface = "bold"))
TYRe <- ComplexHeatmap::Heatmap(t(TYReucMat), cluster_columns = FALSE, column_names_gp = gpar(fontsize = 8), name = "TYR Group", column_title = "TYR Group",  column_title_gp = gpar(fontsize = 20, fontface = "bold"))

hliste <- MYCe+ SHHe + TYRe

ComplexHeatmap::draw(hliste)
```

## All Groups Together

```{r allGroups, echo=FALSE}
library(ComplexHeatmap, quietly = TRUE)
library(BrainSABER, quietly = TRUE)
data("ATRTcosMat","ATRTeucMat")
ComplexHeatmap::Heatmap(t(ATRTcosMat), cluster_columns = FALSE, row_names_gp = gpar(fontsize = 8), name = "Cosine Similarity", column_title = "Cosine Similarity", column_title_gp = gpar(fontsize = 20, fontface = "bold"))

ComplexHeatmap::Heatmap(t(ATRTeucMat), cluster_columns = FALSE, row_names_gp = gpar(fontsize = 8), name = "Euclidean Distance", column_title = "Euclidean Distance", column_title_gp = gpar(fontsize = 20, fontface = "bold"))
