---
title: "Installing and Using BrainSABER"
author: "Carrie Minette"
date: "`r Sys.Date()`"
output:
  html_document: BiocStyle::html_document
  pdf_document: default
vignette: "%\\VignetteIndexEntry{BrainSABER} \n %\\VignetteEncoding{UTF-8} \n %\\VignetteEngine{knitr::rmarkdown}\n"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The Allen Institute for Brain Science provides an 
RNA sequencing (RNA-Seq) data resource for studying 
transcriptional mechanisms involved in human brain 
development known as BrainSpan. This resource serves as an 
additional control in research involving RNA sequencing of 
human brain tissue. BrainSABER facilitates comparisons of user data
with the various developmental stages and brain structures found in
the BrainSpan atlas through an R Shiny interface. BrainSABER utilizes the Bioconductor
ExpressionSet class as a self-validating container for user data
and produces similarity matrices containing samples of the user’s data
analyzed against each sample in the BrainSpan database. These matrices 
are presented as dynamic heat maps, and include both cosine and Euclidean 
similarity methods.

# Installing BrainSABER

BrainSABER requires Bioconductor 3.11 or higher.  For more information on Bioconductor, please see their website at https://bioconductor.org.  To install Bioconductor and BrainSABER, run the following commands in R or RStudio:

```{r, eval=FALSE}
# for R version >= 3.5
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("BrainSABER")
```


After installing, attach the BrainSABER package with the following command:

```{r loadpkg, message=FALSE, warning=FALSE, results='hide'}
library(BrainSABER)
```

# Dynamically Constructing the AIBSARNA Reference Dataset

To leverage the Allen Institute for Brain Science Brain Atlas RNA sequence 
data, BrainSABER includes a function, buildAIBSARNA, that will download the 
data, format it into an ExpressionSet, and use biomaRt to add RefSeq 
identifiers.  This function requires an active internet connection.  

```{r, buildAIB}
AIBSARNA <- buildAIBSARNA()
```

Once AIBSARNA has been build, this function does not need to be run again for 
subsequent work unless the AIBSARNA variable is removed or a new R session is 
started.  If desired, AIBSARNA can be saved as an R data object and reloaded 
for future work without the need to rebuild it.

# The BrainSABER Workflow (R Console)

The BrainSABER workflow can be run either through the R console, or through a Shiny app discussed further below.  In the R console, BrainSABER can be used to compare the Allen Institute data to a Biobase ExpressionSet.  

## Preparing an ExpressionSet object

Biobase ExpressionSet class provides a common interface familiar to those who have analyzed microarray experiments with Bioconductor. The class requires the following three input files:

1. AssayData, a numeric matrix of expression values, where rows are genes, and columns are cells/samples

2. phenoData, an AnnotatedDataFrame object, where rows are cells/samples, and columns are cell/sample attributes (such as cell type, culture condition, day captured, etc.)

3. featureData, an AnnotatedDataFrame object, where rows are features (e.g. genes), and columns are gene attributes, such as gene identifiers, gc content, etc.

The expression value matrix must have the same number of columns as the phenoData has rows, and it must have the same number of rows as the featureData data frame has rows. Row names of the phenoData object should match the column names of the expression matrix. Row names of the featureData object should match row names of the expression matrix.  

For this vignette, we will construct a toy ExpressionSet object by extracting 50 samples and 200 genes from AIBSARNA, using the code below:

```{r getSampleData, message=FALSE, warning=FALSE, results='hide'}
# Obtain the sample indexes to use for subsetting (not random)
sample_idx <- 1:50 * 10 - 1
# Set the RNG seed for repeatable results
set.seed(8)
# Get the total number of genes available
totalGenes <- nrow(AIBSARNA)
# Sample the indices of 200 random genes
gene_idx <- sample.int(totalGenes, 200)

# Subset AIBSARNA
toy_exprs <- exprs(AIBSARNA)[gene_idx, sample_idx]
toy_fd <- fData(AIBSARNA)[gene_idx, ]
toy_pd <- pData(AIBSARNA)[sample_idx, ]

# Convert feature and pheno data to AnnotatedDataFrame required by ExpressionSet

toy_featureData <- AnnotatedDataFrame(data = toy_fd)
toy_phenoData <- AnnotatedDataFrame(data = toy_pd)

# Create toy ExpressionSet
toySet <- ExpressionSet(assayData = toy_exprs,
                        featureData = toy_featureData,
                        phenoData = toy_phenoData)

```

## Select Gene Identifiers

Before we can compare our data to the Allen Institute data, we must remove any genes that are not present in the Allen Institute data.  To accomplish that, we need to explore different combinations of gene identifiers for our data and the Allen Institute data to find which combination yields the most matched genes.  We can use BrainSABER's "getExternalVector()" function to extract trimmed vectors from our toySet, and compare lengths to find which one contains the most matches.

```{r selectid1}
# Try comparing different identifiers
length(getExternalVector(toySet, index = 1, AIBSARNA = AIBSARNA, 
                         dataSetId = "gene_id", AIBSARNAid = "gene_id"))
```
```{r selectid2}
length(getExternalVector(toySet, index = 1, AIBSARNA = AIBSARNA, 
                         dataSetId = "ensembl_gene_id", 
                         AIBSARNAid = "ensembl_gene_id"))
```
```{r selectid3}
length(getExternalVector(toySet, index = 1, AIBSARNA = AIBSARNA, 
                         dataSetId = "gene_symbol", 
                         AIBSARNAid = "gene_symbol"))
```
```{r selectid4}
length(getExternalVector(toySet, index = 1, AIBSARNA = AIBSARNA, 
                         dataSetId = "entrez_id", AIBSARNAid = "entrez_id"))
```
```{r selectid5}
length(getExternalVector(toySet, index = 1, AIBSARNA = AIBSARNA, 
                         dataSetId = "refseq_ids", AIBSARNAid = "refseq_ids"))

```


Because the Allen Institute data was originally sequenced using Ensembl identifiers, the "ensembl_id" column of AIBSARNA is the only column with no missing values, making it the ideal choice if the user data also contains ensembl identifiers.  If not, the Allen Institute data contains other identifiers.  More information is available at their website, http://brainspan.org.

### Filter the Data

Before we can compare our data to the Allen Institute data, we must remove any genes that are not present in the Allen Institute data.  BrainSABER includes a function to trim down user data, which returns a new ExpressionSet.  We will use the "ensembl_gene_id" columns to trim by, as they showed the most matches in the previous step.

```{r filterData}
trimmed_toySet <- getTrimmedExternalSet(dataSet = toySet, 
                                        dataSetId = "ensembl_gene_id",
                                        AIBSARNA = AIBSARNA, 
                                        AIBSARNAid = "ensembl_gene_id")
```

We must also filter AIBSARNA to obtain only the genes present in our data set, using BrainSABER's "getRelevantGenes()" function.  Again, we will use the "ensembl_gene_id" columns.

```{r getRelevantGenes}
trimmed_AIBSARNA <- getRelevantGenes(data = toySet, 
                                     dataSetId = "ensembl_gene_id",
                                     AIBSARNA = AIBSARNA,
                                     AIBSARNAid = "ensembl_gene_id")
```

### Generate Similarity Scores

Once we have filtered the data to contain only genes present in both the user data and the Allen Institute data, we can obtain a data frame containing similarity scores comparing each sample in the user data to each sample in the Allen Institute data.  The getSimScores() function currently supports scoring based on either euclidean distance or cosine similarity.  The scores range from 0 to 1, where 1 is a perfect match, and 0 is completely dissimilar.

```{r getSimScores}
euc_sim <- getSimScores(data = trimmed_toySet, 
                         relevantGenes = trimmed_AIBSARNA,
                         similarity_method = "euclidean")
cos_sim <- getSimScores(data = trimmed_toySet, 
                         relevantGenes = trimmed_AIBSARNA,
                         similarity_method = "cosine")
```

### Generate Age x Structure Matrices or Data Frames

We can use the similarity scores generated in the previous step to generate a matrix or data frame for each sample in our data to identify which ages and brain structures from the Allen Institute data that sample most closely matches.  The getSimMatrix function will return single list which contains the matrices for all samples, and the getSimDataFrame function will return a single list containing data frames for all samples, sorted by the highest score.

```{r getAS}
euc_mats <- getSimMatrix(sim_score = euc_sim, relevantGenes = trimmed_AIBSARNA)
euc_df <- getSimDataFrame(sim_score = euc_sim, 
                          relevantGenes = trimmed_AIBSARNA,
                          similarity_method = "euclidean")
cos_mats <- getSimMatrix(sim_score = cos_sim, relevantGenes = trimmed_AIBSARNA)
cos_df <- getSimDataFrame(sim_score = cos_sim, 
                          relevantGenes = trimmed_AIBSARNA,
                          similarity_method = "cosine")
```

Using R's head() function, we can see the best matches for a sample in our list of data frames:
```{r viewsimdf}
head(cos_df[[1]])
```

The matrices can be fed into a heatmapping function to produce graphics displaying the similarity for a sample.  we will use the heatmaply package to generate an interactive heatmap:

```{r genHeatmap}
library(heatmaply)
heatmaply(euc_mats[[1]])
```

### Generate Up-Regulated / Down-Regulated / Normal Matrices

In addition to similarity scoring, BrainSABER can also be used to create a list of matrices, one for each sample, indicating whether genes are up-reglated, down-regulated, or normal as compared to the Allen Institute data.  The getUNDmatrix() function compares the two data sets, either directly with the "discrete" option, or using the log2 fold-change via "log2fc", and returns either a numerical  matrix or a character matrix where '-1' or 'D' indicates down-regulation, '1' or 'U' indicates up-regulation, and '0' or 'N' indicates normal.  The rows of these matrices will be named with the rownames of the user's trimmed gene expression matrix, and so correspond to the genes of the user data, while the columns of the matrices will be named for the columns in the trimmed Allen Institute data gene expression matrix, and thus correspond to each sample in the Allen Institute data

```{r getundnum}
und_num <- getUNDmatrix(dataSet = trimmed_toySet, 
                        relevantGenes = trimmed_AIBSARNA, 
                        method = "log2fc",
                        matrix_type = "num")
und_num[[1]][1:10, 1:10]
```
```{r getundchar}
und_char <- getUNDmatrix(dataSet = trimmed_toySet, 
                        relevantGenes = trimmed_AIBSARNA, 
                        method = "log2fc",
                        matrix_type = "char")
und_char[[1]][1:10, 1:10]
```


# The BrainSABER Workflow (Shiny app)

BrainSABER comes with a Shiny app that runs most of the BrainSABER workflow, and also generates heatmaps based on the similarity matrices produced by BrainSABER.  This Shiny app requires three .CSV files containing the data required to construct an ExpressionSet.  We can save the toy data we subsetted earlier as .CSV files to use with the Shiny app.  The following code will save these files in your current working directory.

```{r saveCSV}
write.csv(toy_exprs, file = "toy_expression_matrix.csv")
write.csv(as.data.frame(toy_fd), file = "toy_gene_identification_data.csv")
write.csv(as.data.frame(toy_pd), file = "toy_sample_information_data.csv")
```


Once these files are saved, we can run the Shiny app.  The app will guide us through the BrainSABER workflow.
```{r runShinyBrainsABER, eval=FALSE}
runShinyBrainSABER()
```

## Session Info
```{r sessionInfo}
sessionInfo()
```